---
title: "Bayesian AMMIT"
author: "Alessandra Lemos"
#date: "2/16/2022"
output: 
  pdf_document:
    number_sections: true
---

```{r, message=FALSE}
library(bammit)
library(R2jags)
library(ggplot2)
library(gridExtra)
```

# - AMMI Model 
The model is formulated as
$$y_{ij} = \mu + g_i + e_j + \sum\limits_{q = 1}^Q \lambda_q\gamma_{iq}\delta_{jq} + \epsilon_{ij},$$
where $\epsilon_{ij}\sim N(0,\sigma^2_y)$. We have two formulations:


|   Parameter   |                    Josse's                    |               Crossa's              |
|:-------------:|:---------------------------------------------:|:-----------------------------------:|
| $\mu$         |           $N(\mu_mu, \sigma^2_\mu)$           |      $N(\mu_mu, \sigma^2_\mu)$      |
| $g_i$         |               $N(0,\sigma^2_g)$               |        $N(\mu_g,\sigma^2_g)$        |
| $e_j$         |               $N(0,\sigma^2_e)$               |        $N(\mu_e,\sigma^2_e)$        |
| $\lambda_q$   |           $N^+(0,\sigma^2_\lambda)$           | $N^+(\mu_\lambda,\sigma^2_\lambda)$ |
| $\gamma_{iq}$ | $N^+(0,1)$, i = 1; $N(0,1)$, $i>1$, q = 1,...,Q |          spherical uniform          |
| $\delta_{jq}$ |                    $N(0,1)$                   |          spherical uniform          |
| $\sigma^2_y$  |             $U(0,s^2_{\sigma^2_y})$             |       $Inv-Scaled-\chi^2(a,b)$       |
|               |                                                 |                                   |

For both approaches , $\lambda_q>0$ and $\lambda_{q-1}\geq\lambda_q$. In our formulation, we consider $\sigma^2_y \sim \Gamma(a,b)$. In order to meet the model constraints, we perform the following procedure on the bilinear term. Let $\theta^{\gamma}_{iq} \sim N(0,\sigma^\theta)$, for $i = 1, ..., I-1$, and $\theta^{\gamma}_{Iq} = -\sum\limits_{i\neq I}\theta^{\gamma}_{iq}$. Similarly, let  $\theta^{\delta}_{jq} \sim N(0,\sigma^\theta)$, for $j = 1, ..., J-1$, and $\theta^{\delta}_{Jq} = -\sum\limits_{j\neq J}\theta^{\delta}_{jq}$. Then,

$$\gamma_{iq} = \frac{\theta^\gamma_{iq}}{\sqrt{\sum\limits_{i} \theta^{\gamma^2}_{iq}}} \text{ and } \delta_{jq} = \frac{\theta^\delta_{jq}}{\sqrt{\sum\limits_{j} \theta^{\delta^2}_{jq}}}$$
## -  Simulation

Simulation scenarios

* Set $I =\{6, 12\}$, $I =\{4, 6\}$, $Q =\{1,2,3\}$ . 
* Fix $\lambda = \{10, 12, 25\}$.
* Fix $\mu = 100$, $\sigma_g = 10$, $\sigma_e = 10$, $\sigma_y = 2$.

In Jags model:

* Fix $\sigma^2_\theta = 1$.
* Fix $\sigma^2_\theta = 100$.
* Fix $\mu_\lambda = 10$, $\sigma^2_\lambda = 1$, $\mu_g = 0$, $\sigma^2_g = 10$, $\mu_e = 0$, $\sigma^2_e = 10$, $a = 0.1$, $b = 0.1$. 

## Jags implementation
```{r}
 modelCode <- "
  model
  {
  # Likelihood
   for (k in 1:N) {
    Y[k] ~ dnorm(mu[k], sy^-2)
    mu[k] = muall + g[genotype[k]] + e[environment[k]] + blin[k]
    blin[k] = sum(lambda[1:Q] * gamma[genotype[k],1:Q] * delta[environment[k],1:Q])
   }

  # Priors
  # Prior on grand mean
   muall ~ dnorm(mmu, smu^-2)

  # Prior on genotype effect
  for(i in 1:I) {
  g[i] ~ dnorm(mug, sg^-2) # Prior on genotype effect
  }

  # Prior on environment effect
  for(j in 1:J) {
  e[j] ~ dnorm(mue, se^-2) # Prior on environment effect
  }


  # Priors on gamma
  for(q in 1:Q){
    for(i in 1:(I-1)){
      theta[i,q] ~ dnorm(0,stheta)
    }
    theta[I,q] = -sum(theta[1:(I-1),q])
    thetaSum[q] = sqrt(sum(theta[1:I,q]^2)) + 0.00001
    for(i in 1:I){
      gamma[i,q] = theta[i,q]/thetaSum[q]
    }
  }

  # Priors on delta
   for(q in 1:Q){
    for(j in 1:(J-1)){
      aux[j,q] ~ dnorm(0,stheta)
    }
    aux[J,q] = -sum(aux[1:(J-1),q])
    auxSum[q] = sqrt(sum(aux[1:J,q]^2)) + 0.000001
    for(j in 1:J){
      delta[j,q] = aux[j,q]/auxSum[q]
    }
  }

  # Prior on eigenvalues
  for(q in 1:Q) {
    lambda_raw[q] ~ dnorm(mulambda, slambda^-2)T(0,)
  }
  lambda = sort(lambda_raw)

  # Prior on residual standard deviation
   sy ~ dgamma(a, b) # inverse of sy
  }
  "
```


## - Results

### $\sigma^2_\theta = 1$

#### Q = 1

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
data <- simulateDataAmmi(I = 12, J = 6, mu = 100, sg = 10, se = 10, sy = 2, lambda = 10)
josseModel <- josseJags(data = data,  mmu = 90, smu = 10,
                        sg = 10, se = 10, slambda = 1,
                        a = 0.1, b = 0.1, nthin = 1, nburnin = 100)
crossaModel <- crossaJags(data = data, mmu = 90, smu = 10, 
                          mug = 0, sg = 10, mue = 0, se = 10, 
                          mulambda = 10, slambda = 1, a = 0.1, b = 0.1, stheta = 1,
                          nthin = 1, nburnin = 100)
grid.arrange(
qplot(data$blin, josseModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Josse", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
qplot(data$blin, crossaModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Crossa", xlab = "True bilinear") + geom_abline() +  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
nrow = 1)
```


### Q = 2

```{r, echo=FALSE, message=FALSE, warning = FALSE, results='hide'}
data <- simulateDataAmmi(I = 12, J = 6, mu = 100, sg = 10, se = 10, sy = 2, lambda = c(10,12))
josseModel <- josseJags(data = data,  mmu = 90, smu = 10,
                        sg = 10, se = 10, slambda = 1,
                        a = 0.1, b = 0.1, nthin = 1, nburnin = 100)
crossaModel <- crossaJags(data = data, mmu = 90, smu = 10, 
                          mug = 0, sg = 10, mue = 0, se = 10, 
                          mulambda = 10, slambda = 1, a = 0.1, b = 0.1, stheta = 1,
                          nthin = 1, nburnin = 100)
grid.arrange(
qplot(data$blin, josseModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Josse", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
qplot(data$blin, crossaModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Crossa", xlab = "True bilinear") + geom_abline() +  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
nrow = 1)
```
```{r, echo=FALSE}
yPredAmmi <- predictionAMMI(crossaModel, data)
```
```{r, echo=FALSE}
g <- qplot(data$g, crossaModel$BUGSoutput$mean$g, ylab = expression(hat(g)), xlab = "g") + geom_abline() +
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
e <- qplot(data$e, crossaModel$BUGSoutput$mean$e , ylab = expression(hat(e)), xlab = "e") + geom_abline()+  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
yhat <- qplot(yPredAmmi, data$y, xlab = expression(hat(y)), ylab = "y") + geom_abline() + 
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
grid.arrange(g,e,yhat, nrow = 1)
```

### Q = 3

```{r, echo=FALSE, message=FALSE, warning = FALSE, results='hide'}
data <- simulateDataAmmi(I = 12, J = 6, mu = 100, sg = 10, se = 10, sy = 2, lambda = c(10,12, 25))
josseModel <- josseJags(data = data,  mmu = 90, smu = 10,
                        sg = 10, se = 10, slambda = 1,
                        a = 0.1, b = 0.1, nthin = 1, nburnin = 100)
crossaModel <- crossaJags(data = data, mmu = 90, smu = 10, 
                          mug = 0, sg = 10, mue = 0, se = 10, 
                          mulambda = 10, slambda = 1, a = 0.1, b = 0.1, stheta = 1,
                          nthin = 1, nburnin = 100)
grid.arrange(
qplot(data$blin, josseModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Josse", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
qplot(data$blin, crossaModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Crossa", xlab = "True bilinear") + geom_abline() +  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
nrow = 1)
```
```{r, echo=FALSE}
yPredAmmi <- predictionAMMI(crossaModel, data)
```
```{r, echo=FALSE}
g <- qplot(data$g, crossaModel$BUGSoutput$mean$g, ylab = expression(hat(g)), xlab = "g") + geom_abline() +
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
e <- qplot(data$e, crossaModel$BUGSoutput$mean$e , ylab = expression(hat(e)), xlab = "e") + geom_abline()+  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
yhat <- qplot(yPredAmmi, data$y, xlab = expression(hat(y)), ylab = "y") + geom_abline() + 
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
grid.arrange(g,e,yhat, nrow = 1)
```

### $\sigma^2_\theta = 100$

#### Q = 1

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
data <- simulateDataAmmi(I = 12, J = 6, mu = 100, sg = 10, se = 10, sy = 2, lambda = 10)
josseModel <- josseJags(data = data,  mmu = 90, smu = 10,
                        sg = 10, se = 10, slambda = 1,
                        a = 0.1, b = 0.1, nthin = 1, nburnin = 100)
crossaModel <- crossaJags(data = data, mmu = 90, smu = 10, 
                          mug = 0, sg = 10, mue = 0, se = 10, 
                          mulambda = 10, slambda = 1, a = 0.1, b = 0.1, stheta = 100,
                          nthin = 1, nburnin = 100)
grid.arrange(
qplot(data$blin, josseModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Josse", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
qplot(data$blin, crossaModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Crossa", xlab = "True bilinear") + geom_abline() +  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
nrow = 1)
```


### Q = 2

```{r,echo=FALSE,message=FALSE, warning=FALSE, results='hide'}
data <- simulateDataAmmi(I = 12, J = 6, mu = 100, sg = 10, se = 10, sy = 2, lambda = c(10, 12))
josseModel <- josseJags(data = data,  mmu = 90, smu = 10,
                        sg = 10, se = 10, slambda = 1,
                        a = 0.1, b = 0.1, nthin = 1, nburnin = 100)
crossaModel <- crossaJags(data = data, mmu = 90, smu = 10, 
                          mug = 0, sg = 10, mue = 0, se = 10, 
                          mulambda = 10, slambda = 1, a = 0.1, b = 0.1, stheta = 100,
                          nthin = 1, nburnin = 100)
grid.arrange(qplot(data$blin, josseModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Josse", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
             qplot(data$blin, crossaModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Crossa", xlab = "True bilinear") + geom_abline() +  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
             nrow = 1)
```

```{r, echo=FALSE}
yPredAmmi <- predictionAMMI(crossaModel, data)
```

```{r, echo=FALSE}
g <- qplot(data$g, crossaModel$BUGSoutput$mean$g, ylab = expression(hat(g)), xlab = "g") + geom_abline() +
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
e <- qplot(data$e, crossaModel$BUGSoutput$mean$e , ylab = expression(hat(e)), xlab = "e") + geom_abline()+  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
yhat <- qplot(yPredAmmi, data$y, xlab = expression(hat(y)), ylab = "y") + geom_abline() + 
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
grid.arrange(g,e,yhat, nrow = 1)
```

### Q = 3

```{r,echo=FALSE,message=FALSE, warning=FALSE, results='hide'}
data <- simulateDataAmmi(I = 12, J = 6, mu = 100, sg = 10, se = 10, sy = 2, lambda = c(10, 12, 25))
josseModel <- josseJags(data = data,  mmu = 90, smu = 10,
                        sg = 10, se = 10, slambda = 1,
                        a = 0.1, b = 0.1, nthin = 1, nburnin = 100)
crossaModel <- crossaJags(data = data, mmu = 90, smu = 10, 
                          mug = 0, sg = 10, mue = 0, se = 10, 
                          mulambda = 10, slambda = 1, a = 0.1, b = 0.1, stheta = 100,
                          nthin = 1, nburnin = 100)
grid.arrange(qplot(data$blin, josseModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Josse", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
             qplot(data$blin, crossaModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - Crossa", xlab = "True bilinear") + geom_abline() +  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6),
             nrow = 1)
```

```{r, echo=FALSE}
yPredAmmi <- predictionAMMI(crossaModel, data)
```

```{r, echo=FALSE}
g <- qplot(data$g, crossaModel$BUGSoutput$mean$g, ylab = expression(hat(g)), xlab = "g") + geom_abline() +
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
e <- qplot(data$e, crossaModel$BUGSoutput$mean$e , ylab = expression(hat(e)), xlab = "e") + geom_abline()+  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
yhat <- qplot(yPredAmmi, data$y, xlab = expression(hat(y)), ylab = "y") + geom_abline() + 
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
grid.arrange(g,e,yhat, nrow = 1)
```


# - BAMMIT Model

The BAMMIT model is formulated as
$$y_{ijk} = \mu + g_i + e_j + t_k + \sum\limits_{q = 1}^Q \lambda_q\gamma_{iq}\delta_{jq}\kappa_{kq} + \epsilon_{ijk},$$

where $\epsilon_{ijk}\sim N(0,\sigma^2_y)$, $i = 1,..., I$, $j = 1, ..., J$ and $k = 1, ..., K$. We Assume that

$$\mu\sim N(\mu_\mu, \sigma^2_\mu)$$      
$$g_i\sim N(\mu_g,\sigma^2_g)$$        
$$e_j\sim N(\mu_e,\sigma^2_e)$$   
$$t_k\sim N(\mu_t,\sigma^2_t)$$   
$$\lambda_q \sim N^+(\mu_\lambda,\sigma^2_\lambda)$$         
$$\sigma^2_y\sim \Gamma(a,b)$$

For the parameters $\gamma_{iq}$, $\delta_{jq}$ and $\kappa_{kq}$ we follow as before. Let $$\theta^{\gamma}_{iq} \sim N(0,\sigma^\theta), i = 1, ..., I-1,$$
$$\theta^{\gamma}_{Iq} = -\sum\limits_{i\neq I}\theta^{\gamma}_{iq}$$
$$\theta^{\delta}_{jq} \sim N(0,\sigma^\theta), j = 1, ..., J-1,$$ 
$$\theta^{\delta}_{Jq} = -\sum\limits_{j\neq J}\theta^{\delta}_{jq};$$
$$\theta^{\kappa}_{kq} \sim N(0,\sigma^\theta), k = 1, ..., K-1,$$ 
$$\theta^{\kappa}_{Kq} = -\sum\limits_{k\neq K}\theta^{\kappa}_{kq};$$

## -  Simulation

Simulation scenarios

* Set $I =\{6, 12\}$, $I =\{4, 6\}$, $K = \{2\}$ $Q =\{1\}$ . 
* Fix $\lambda = \{10\}$.
* Fix $\mu = 100$, $\sigma_g = 10$, $\sigma_e = 10$, $\sigma_t = 10$, $\sigma_y = 2$.

In Jags model:

* Fix $\sigma^2_\theta = 1$.
* Fix $\sigma^2_\theta = 100$.
* Fix $\mu_\lambda = 10$, $\sigma^2_\lambda = 1$, $\mu_g = 0$, $\sigma^2_g = 10$, $\mu_e = 0$, $\sigma^2_e = 10$, $\mu_t = 0$, $\sigma^2_t = 10$, $a = 0.1$, $b = 0.1$. 

### $\sigma^2_\theta = 1$
```{r, echo = FALSE, results='hide'}
dataT <- simulateDataBammit(I = 6, J = 4, K = 2, mu = 100, sg = 10, se = 10, st = 10, sy = 2, lambda = 10)
bammitModel <- bammitJags(data = dataT, mmu = 90, smu = 10, mug = 0, sg = 10, mue = 0, se = 10, mut = 0, st = 10, mulambda = 10, slambda = 1, stheta = 1, a = 0.1, b = 0.1, nthin = 1, nburnin = 100)

qplot(dataT$blin, bammitModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - BAMMIT", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
```

```{r}
yhatB <- predictionBAMMIT(bammitModel, dataT)
qplot(yhatB, dataT$y, xlab = expression(hat(y)), ylab = "y") + geom_abline() + 
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
```


### $\sigma^2_\theta = 100$
```{r, echo = FALSE, results='hide'}
dataT <- simulateDataBammit(I = 6, J = 4, K = 2, mu = 100, sg = 10, se = 10, st = 10, sy = 2, lambda = 10)
bammitModel <- bammitJags(data = dataT, mmu = 90, smu = 10, mug = 0, sg = 10, mue = 0, se = 10, mut = 0, st = 10, mulambda = 10, slambda = 1, stheta = 100, a = 0.1, b = 0.1, nthin = 1, nburnin = 100)

qplot(dataT$blin, bammitModel$BUGSoutput$mean$blin, ylab = "Estimated bilinear - BAMMIT", xlab = "True bilinear") + geom_abline() + geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
```

```{r}
yhatB <- predictionBAMMIT(bammitModel, dataT)
qplot(yhatB, dataT$y, xlab = expression(hat(y)), ylab = "y") + geom_abline() + 
  geom_abline() + theme_bw() + geom_point(colour = "red", size = 0.6)
```









